---
title: "Delta Sharing vNext"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Delta Sharing vNext}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`delta.sharing` uses immutable S7 descriptors for profiles, clients, tables,
snapshot reads, and change data feed (CDF) reads. Constructing these objects is
offline: it validates configuration but does not make an HTTP request or
exchange an OAuth token.

The implementation keeps the user-facing model deliberately small:

1. create a client from a standard Delta Sharing profile;
2. identify a table;
3. describe a snapshot or CDF read; and
4. materialize that descriptor as an Arrow stream, Arrow table, or data frame.

## Create descriptors

The inline profile below contains an explicitly non-secret example value. In
an application, pass `sharing_client()` the path to a protected profile file
instead.

```{r}
library(delta.sharing)

profile <- sharing_profile(list(
  shareCredentialsVersion = 2,
  type = "bearer_token",
  endpoint = "https://sharing.example.test/api",
  bearerToken = "example-only-not-a-secret"
))
client <- sharing_client(profile)

orders <- sharing_table(client, "sales.default.orders")
orders
```

A compact table name is unambiguous only when its three components contain no
dots. Supply the components separately when a name contains a dot:

```{r}
events <- sharing_table(
  client,
  share = "product",
  schema = "default",
  table = "events.v2"
)
table_identifier(events)@table
```

The descriptors expose read-only S7 properties. Creating a read returns a new
object and does not modify the reusable table:

```{r}
latest_orders <- sharing_read(
  orders,
  columns = c("order_id", "ordered_at", "amount"),
  limit = 1000,
  predicate = list(
    op = "equal",
    column = "region",
    value = "APAC"
  )
)

orders_at_version <- sharing_read(orders, version = 42)
orders_at_time <- sharing_read(
  orders,
  timestamp = as.POSIXct("2026-01-01 00:00:00", tz = "UTC")
)

latest_orders@columns
orders_at_version@version
orders_at_time@timestamp
```

`limit` is enforced exactly by the Kernel scan. A structured `predicate` is a
best-effort server hint, not an exact row filter.

Snapshot requests accept `response_format = "auto"`, `"delta"`, or
`"parquet"`. Both Delta- and Parquet-format responses ultimately use the same
Delta Kernel Arrow stream. For Parquet responses, R validates and normalizes
the protocol actions into a private Kernel-readable log; the package does not
use a second Parquet reader.

Reader capabilities are response-specific. Delta snapshot requests advertise
`columnmapping` and `timestampntz`; Parquet-response normalization does not
advertise Delta reader features. `deletionvectors` remains intentionally
unadvertised until exact absolute-path HTTPS deletion-vector resolution has
end-to-end proof.

## Describe change data

CDF is a separate immutable descriptor. Materialization currently requires a
Delta-format response and explicit inclusive start and end versions:

```{r}
order_changes <- sharing_changes(
  orders,
  starting_version = 120,
  ending_version = 125,
  columns = c(
    "order_id",
    "status",
    "_change_type",
    "_commit_version",
    "_commit_timestamp"
  ),
  response_format = "delta"
)
order_changes
```

Open-ended CDF, timestamp-bound CDF, and Parquet-format CDF are not yet
materialized. They fail with a typed unsupported condition before HTTP so the
package never guesses provider versions.

## Discover and inspect remote data

The discovery and metadata generics perform authenticated network calls. With
a real profile, use them to find data and inspect a table without scanning its
rows:

```{r, eval = FALSE}
client <- sharing_client("/secure/path/to/config.share")

list_shares(client)
list_schemas(client, share = "sales")
list_tables(client, share = "sales", schema = "default")

orders <- sharing_table(client, "sales.default.orders")
table_version(orders)
table_protocol(orders)
table_metadata(orders)
table_schema(orders)
```

These results are safe projections. They do not expose credentials, signed
storage URLs, response bodies, or private paths.

## Materialize one reader three ways

`read_arrow_stream()` is the primary materializer. It returns a lazy,
single-consumer `nanoarrow_array_stream` whose batches are bounded by
`batch_size`. Release the stream explicitly if consumption stops early:

```{r, eval = FALSE}
stream <- read_arrow_stream(latest_orders, batch_size = 65536L)
diagnostics <- read_diagnostics(stream)

# Pull batches using nanoarrow or pass the stream to an Arrow-aware consumer.

stream$release()

# Diagnostics are immutable and remain readable after release.
read_diagnostics(stream)
```

Normal exhaustion, explicit release, and finalization release the native
Kernel scan and its private temporary log. An R interrupt during a stream pull
or eager materialization cancels and releases the native stream, then raises a
typed `delta_sharing_cancelled` condition on the owning R thread.

The eager adapters consume exactly that same stream:

```{r, eval = FALSE}
orders_df <- read_data_frame(latest_orders)
orders_arrow <- read_arrow(latest_orders) # requires the optional arrow package
```

`read_data_frame()` allocates the complete result in R memory. `read_arrow()`
imports the Arrow C Stream directly and fails before HTTP or Kernel work when
the optional `arrow` package is unavailable.

`read_diagnostics()` returns stream-local, redacted planning facts such as the
selected format, resolved version or CDF bounds, page and file counts,
projection, limit, batch size, and signed-URL expiry summary. It never contains
credentials, URLs, paths, tokens, predicate values, protocol actions, or
mutable execution state.

## Ownership boundary and current limits

R owns profiles, authentication, HTTP, retries, protocol parsing, request
planning, Parquet-response normalization, synthetic logs, public conditions,
diagnostics, and eager adapters. Rust is limited to Delta Kernel scan
execution, Arrow C Stream export, panic containment, cancellation, and the
native resources whose lifetime is coupled to the stream.

Two public limitations are intentional:

- `read_schema()` does not yet return a projected read schema. Use
  `table_schema()` for the table's logical schema.
- `batch_size` is supported, but any non-`NULL` `concurrency` argument is
  rejected before materialization.
- Comparable local performance and lifecycle evidence exists and keeps this
  native boundary narrow, but release and target-platform performance gates
  remain unresolved.

## Serialization is intentionally inert

Descriptors are safe to serialize because credential and token state is not an
S7 property. A deserialized profile, client, table, or read is deliberately
inert: it cannot recover the original process-local authentication handle.
Reconstruct the client from its protected profile source in the receiving
process before making requests.

```{r}
raw_client <- serialize(client, NULL)
restored_client <- unserialize(raw_client)

# Safe descriptor metadata remains available.
restored_client@profile@endpoint
```
