---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`delta.sharing` is an R6 client for [Delta Sharing](https://delta.io/sharing/).
R manages profiles, authentication, discovery, and query planning. Delta Kernel
reads snapshot and change data feed rows and exposes them through an Arrow C
Stream.

Most workflows have four steps:

1. create a client from a standard Delta Sharing profile;
2. get a reusable table handle;
3. configure a snapshot or change data feed read; and
4. materialize it as an Arrow stream, an Arrow table, or a data frame.

## Connect to a share

### Try the open dataset server

The Delta Sharing project hosts an open server with example tables. Its
[public profile](https://github.com/delta-io/delta-sharing/blob/main/examples/open-datasets.share)
can be used without registering or creating a private credential:

```{r, eval = FALSE}
library(delta.sharing)

profile_url <- paste0(
  "https://raw.githubusercontent.com/",
  "delta-io/delta-sharing/main/examples/open-datasets.share"
)

open_profile <- httr2::request(profile_url) |>
  httr2::req_perform() |>
  httr2::resp_body_string()

client <- sharing_client(open_profile)
client$list_tables()
```

Read a few rows from one of the public tables:

```{r, eval = FALSE}
housing <- client$table("delta_sharing.default.boston-housing")
housing$snapshot(limit = 5)$to_data_frame()
```

### Use your own profile

```{r, eval = FALSE}
library(delta.sharing)

client <- sharing_client("~/config.share")
```

`sharing_client()` accepts:

- a path to a `.share` file;
- an inline JSON string; or
- a parsed list.

Profile versions 1 and 2 support bearer tokens, basic auth, OAuth client
credentials, and private-key JWT client assertions.

Creating the client is offline: it parses the profile without making a request
or exchanging a token. When authentication is first needed, `httr2` handles
token exchange, caching, and refresh.

## Discover data

```{r, eval = FALSE}
client$list_shares()
client$list_schemas()
client$list_tables()
```

Discovery follows every page and returns tibbles. Supply `share` or both
`share` and `schema` only when you want to narrow the result.

## Inspect a table

A table handle is cheap and reusable. Metadata queries do not scan rows.

```{r, eval = FALSE}
orders <- client$table("sales.default.orders")

orders$version()
orders$protocol()
orders$metadata()
orders$schema()
```

Use explicit components when a name contains a dot:

```{r, eval = FALSE}
events <- client$table(
  share = "product",
  schema = "default",
  name = "events.v2"
)
```

## Read a snapshot

Query options live on `snapshot()`; the returned object exposes materializers.

```{r, eval = FALSE}
snapshot <- orders$snapshot(
  columns = c("order_id", "amount"),
  limit = 1000
)

# Lazy Arrow C stream (use when the result may not fit in memory).
stream <- snapshot$to_arrow_stream(batch_size = 65536L)

# Lazy Arrow reader for consumers such as DuckDB.
reader <- snapshot$to_arrow_reader()
reader$Close()

# Eager adapters over the same stream path.
df <- snapshot$to_data_frame()
tbl <- snapshot$to_arrow() # requires the optional {arrow} package
```

Read a specific version or point in time:

```{r, eval = FALSE}
orders$snapshot(version = 42)$to_data_frame()
orders$snapshot(timestamp = as.POSIXct("2024-01-01", tz = "UTC"))$to_data_frame()
orders$snapshot(timestamp = "2024-01-01T00:00:00Z")$to_data_frame()
```

`limit` is enforced exactly by the Kernel scan. A `predicate` is a best-effort
server hint, not an exact row filter.

The default `response_format = "auto"` is negotiated once per table within a
client and reused by later reads. Explicit `"delta"` or `"parquet"` values
bypass negotiation. Table metadata and schema methods continue to fetch their
current values.

Predicates use ordinary nested R lists. The package serializes the list to the
JSON-string wire field required by Delta Sharing:

```{r, eval = FALSE}
active_orders <- list(
  op = "equal",
  children = list(
    list(op = "column", name = "is_active", valueType = "bool"),
    list(op = "literal", value = "true", valueType = "bool")
  )
)
orders$snapshot(predicate = active_orders)$to_data_frame()
```

## Read a change data feed

```{r, eval = FALSE}
orders$changes(starting_version = 120, ending_version = 125)$to_data_frame()
orders$changes(
  starting_timestamp = "2024-01-01T00:00:00Z",
  ending_timestamp = "2024-01-02T00:00:00Z"
)$to_data_frame()
```

## How reads work

A read follows one path:

1. R sends the query to the Delta Sharing server.
2. R writes the response to a private, temporary Delta log.
3. Delta Kernel reads that log, fetches the data files directly from their
   presigned URLs, and exposes the rows as an Arrow stream.

Delta and Parquet responses both use this path. R does not download the data
files separately.
