---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`delta.sharing` is an R6 client for [Delta Sharing](https://delta.io/sharing/).
R manages profiles, authentication, discovery, and query planning. Delta Kernel
reads snapshot and change data feed rows and exposes them through an Arrow C
Stream.

Most workflows have four steps:

1. create a client from a standard Delta Sharing profile;
2. get a reusable table handle;
3. configure a snapshot or change data feed read; and
4. materialize it as an Arrow stream, an Arrow table, or a data frame.

## Connect to a share

### Try the open dataset server

The Delta Sharing project hosts an open server with example tables. Its
[public profile](https://github.com/delta-io/delta-sharing/blob/main/examples/open-datasets.share)
can be used without registering or creating a private credential:

```{r, eval = FALSE}
library(delta.sharing)

client <- sharing_client(demo_profile())
client$list_tables("delta_sharing")
```

Read a few rows from one of the public tables:

```{r, eval = FALSE}
housing <- client$table("delta_sharing.default.boston-housing")
housing$snapshot(limit = 5)$to_data_frame()
```

### Use your own profile

```{r, eval = FALSE}
library(delta.sharing)

client <- sharing_client("~/config.share")
```

`sharing_client()` accepts:

- a path to a `.share` file; or
- a parsed list.

Profile versions 1 and 2 support bearer tokens, basic auth, OAuth client
credentials, and private-key JWT bearer authentication.

Creating the client is offline: it parses the profile without making a request
or exchanging a token. When authentication is first needed, `httr2` handles
token exchange, caching, and refresh.

## Discover data

```{r, eval = FALSE}
client$list_shares()
client$list_schemas("delta_sharing")
client$list_tables("delta_sharing")
```

Discovery follows every page and returns printable lists of records. Listing
schemas or tables requires a share; supply a schema to narrow a table listing.

## Inspect a table

A table handle is cheap and reusable. Metadata queries do not scan rows.

```{r, eval = FALSE}
orders <- client$table("sales.default.orders")

orders$version()
orders$protocol()
orders$metadata()
orders$schema()
```

Use explicit components when a name contains a dot:

```{r, eval = FALSE}
events <- client$table(
  share = "product",
  schema = "default",
  name = "events.v2"
)
```

## Read a snapshot

Query options live on `snapshot()`; the returned object exposes materializers.

```{r, eval = FALSE}
snapshot <- orders$snapshot(
  columns = c("order_id", "amount"),
  limit = 1000
)

# Lazy Arrow C stream (use when the result may not fit in memory).
stream <- snapshot$to_arrow_stream(batch_size = 65536L)
# Pass the stream to a compatible consumer, or release it when unused.
stream$release()

# Lazy Arrow reader for consumers such as DuckDB.
reader <- snapshot$to_arrow_reader()
reader$Close()

# Eager adapters over the same stream path.
df <- snapshot$to_data_frame()
tbl <- snapshot$to_arrow() # requires the optional {arrow} package
```

Read a specific version or point in time:

```{r, eval = FALSE}
orders$snapshot(version = 42)$to_data_frame()
orders$snapshot(timestamp = as.POSIXct("2024-01-01", tz = "UTC"))$to_data_frame()
orders$snapshot(timestamp = "2024-01-01T00:00:00Z")$to_data_frame()
```

`limit` is enforced exactly by the Kernel scan. A `predicate` is a best-effort
server hint, not an exact row filter.

The default `response_format = "auto"` is negotiated once per table within a
client and reused by later reads. Explicit `"delta"` or `"parquet"` values
bypass negotiation. Table metadata and schema methods continue to fetch their
current values.

Predicates use ordinary nested R lists. The package serializes the list to the
JSON-string wire field required by Delta Sharing:

```{r, eval = FALSE}
active_orders <- list(
  op = "equal",
  children = list(
    list(op = "column", name = "is_active", valueType = "bool"),
    list(op = "literal", value = "true", valueType = "bool")
  )
)
orders$snapshot(predicate = active_orders)$to_data_frame()
```

## Read a change data feed

```{r, eval = FALSE}
orders$changes(starting_version = 120, ending_version = 125)$to_data_frame()
orders$changes(
  starting_timestamp = "2024-01-01T00:00:00Z",
  ending_timestamp = "2024-01-02T00:00:00Z"
)$to_data_frame()
```

## Reuse downloaded files

Downloads are private to one read by default. For repeated reads of the same
table, opt into the session cache on `snapshot()` or `changes()`:

```{r, eval = FALSE}
snapshot <- orders$snapshot(cache = TRUE)

first <- snapshot$to_data_frame(threads = 4)
second <- snapshot$to_data_frame() # reuses unchanged Delta data files

orders$clear_cache()
```

`threads` controls the maximum number of concurrent downloads and defaults to
four. `clear_cache()` removes only that table's cached downloads; active readers
remain valid. The full session cache is removed when the package unloads.

## How reads work

A read follows one path:

1. R sends the query to the Delta Sharing server.
2. R downloads the selected Parquet and deletion-vector files into a private
   local table, using bounded parallel requests.
3. R writes the response as a synthetic Delta log beside those files.
4. Delta Kernel reads the local table and exposes the rows as an Arrow stream.

Delta and Parquet responses, including change data feeds, use this path. Delta
Kernel remains responsible for projection, deletion vectors, column mapping,
change semantics, and Arrow batches.

R checks for user interrupts between Arrow batches. If a batch pull is already
waiting inside Delta Kernel's local I/O, that pull must return or fail before R
can handle the interrupt. Releasing a stream between pulls still releases the
Kernel reader and its private read directory immediately.
