---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`delta.sharing` connects R to tables exposed through
[Delta Sharing](https://delta.io/sharing/). A typical workflow is to create a
client, select a table, describe a read, and choose the form of the result.

## Try the public example data

The Delta Sharing project hosts an open server that can be used without
registering or creating a private credential:

```{r, eval = FALSE}
library(delta.sharing)

client <- sharing_client(demo_profile())

housing <- client$table("delta_sharing.default.boston-housing")
housing$snapshot(
  columns = c("chas", "medv"),
  limit = 5
)$to_tibble()
```

`demo_profile()` fetches the public profile maintained by the Delta Sharing
project. The same client and table methods work with a private share.

## Connect to your share

Pass `sharing_client()` either a path to a `.share` profile or a parsed profile
list:

```{r, eval = FALSE}
client <- sharing_client("~/config.share")
```

Creating a client parses the profile without making a request or exchanging a
token. Authentication is performed when the first authenticated request is
made.

## Find a table

Use the client to discover the data available through its profile:

```{r, eval = FALSE}
client$list_shares()
client$list_schemas("sales")
client$list_tables("sales", "default")
```

Discovery follows every result page and returns printable lists of records.
Create a reusable table handle from a listed table:

```{r, eval = FALSE}
orders <- client$table("sales.default.orders")

orders
orders$version()
orders$schema()
```

The table handle is cheap to create and does not read table rows. Its
`protocol()` and `metadata()` methods expose additional table details.

## Read a snapshot

Configure the read with `snapshot()`, then materialize it. `to_tibble()` is the
usual choice for R analysis:

```{r, eval = FALSE}
snapshot <- orders$snapshot(
  columns = c("order_id", "status", "amount"),
  limit = 1000
)

orders_tbl <- snapshot$to_tibble()
```

Snapshots can also target a table version or point in time:

```{r, eval = FALSE}
orders$snapshot(version = 42)$to_tibble()
orders$snapshot(timestamp = "2024-01-01T00:00:00Z")$to_tibble()
```

Choose the materializer that matches the next consumer:

| Result | Method |
|---|---|
| Tibble for ordinary R analysis | `to_tibble()` |
| Base data frame | `to_data_frame()` |
| In-memory Arrow table | `to_arrow()` |
| Lazy Arrow reader, including for DuckDB | `to_arrow_reader()` |
| Low-level Arrow C Stream | `to_arrow_stream()` |

The Arrow table and reader methods require the optional `arrow` package. Each
materializer call starts its own read, so choose one rather than calling several
on the same snapshot just to convert the result.

Column projection and `limit` are applied by Delta Kernel. Predicate hints are
expressed as nested R lists and sent to the sharing server as best-effort hints;
they are not exact row filters. See `?SharingTable` for the snapshot options and
predicate structure.

## Read changes

For a table with change data feed enabled, `changes()` reads an inclusive
version or timestamp range:

```{r, eval = FALSE}
changes_tbl <- orders$changes(
  starting_version = 120,
  ending_version = 125,
  columns = c(
    "order_id",
    "status",
    "_change_type",
    "_commit_version",
    "_commit_timestamp"
  )
)$to_tibble()
```

`_change_type` distinguishes inserts, deletes, and the before and after rows of
updates. The commit columns identify when each change was recorded. Change
reads expose the same materializers as snapshots.

## Downloads and caching

Before Delta Kernel reads the rows, the package downloads the files selected by
the sharing server into a session cache. Up to four files are downloaded at a
time by default, and later handles for the same endpoint and table reuse files
already present there. The cache lives in R's session temporary directory, so
it normally requires no user management.

## How a read works

R sends the Delta Sharing request, stages the selected data and
deletion-vector files, and writes a local Delta log that points to them. Delta
Kernel then applies projection, deletion vectors, column mapping, and change
semantics before exposing Arrow batches. The requested materializer turns those
batches into the selected R or Arrow result.

## Next steps

See `?SharingTable`, `?SharingSnapshot`, and `?SharingChanges` for the complete
read options. The
[README](https://github.com/zacdav-db/delta-sharing-r#query-with-duckdb) shows
how to query a shared table with DuckDB without first creating an R data frame.
