---
title: "Delta Sharing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Delta Sharing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`delta.sharing` is an R6 client for [Delta Sharing](https://delta.io/sharing/).
R manages profiles, authentication, the control plane, and query planning;
snapshot and change data feed row scans are read by Delta Kernel and returned
through an Arrow C Stream.

The user-facing model is deliberately small and follows a staged shape:

1. create a client from a standard Delta Sharing profile;
2. get a reusable table handle;
3. configure a snapshot or change data feed read; and
4. materialize it as an Arrow stream, an Arrow table, or a data frame.

## Create a client

```{r, eval = FALSE}
library(delta.sharing)

client <- sharing_client("~/config.share")
```

The profile may be a path to a `.share` file, a parsed list, or an inline JSON
string. Profile versions 1 and 2 are supported, including bearer tokens, basic
auth, OAuth client credentials, and private-key JWT client assertions.
Construction is offline: it parses configuration but makes no request and
exchanges no token. Token exchange, caching, and refresh are handled by `httr2`
on the first authenticated request.

## Discover data

```{r, eval = FALSE}
client$list_shares()
client$list_schemas()
client$list_tables()
```

Discovery follows every page and returns tibbles. Supply `share` or both
`share` and `schema` only when you want to narrow the result.

## Inspect a table

A table handle is cheap and reusable. Metadata queries do not scan rows.

```{r, eval = FALSE}
orders <- client$table("sales.default.orders")

orders$version()
orders$protocol()
orders$metadata()
orders$schema()
```

Use explicit components when a name contains a dot:

```{r, eval = FALSE}
events <- client$table(
  share = "product",
  schema = "default",
  name = "events.v2"
)
```

## Read a snapshot

Query options live on `snapshot()`; the returned object exposes materializers.

```{r, eval = FALSE}
snapshot <- orders$snapshot(
  columns = c("order_id", "amount"),
  limit = 1000
)

# Lazy Arrow C stream (use when the result may not fit in memory).
stream <- snapshot$to_arrow_stream(batch_size = 65536L)

# Lazy Arrow reader for consumers such as DuckDB.
reader <- snapshot$to_arrow_reader()
reader$Close()

# Eager adapters over the same stream path.
df <- snapshot$to_data_frame()
tbl <- snapshot$to_arrow() # requires the optional {arrow} package
```

Read a specific version or point in time:

```{r, eval = FALSE}
orders$snapshot(version = 42)$to_data_frame()
orders$snapshot(timestamp = as.POSIXct("2024-01-01", tz = "UTC"))$to_data_frame()
orders$snapshot(timestamp = "2024-01-01T00:00:00Z")$to_data_frame()
```

`limit` is enforced exactly by the Kernel scan. A `predicate` is a best-effort
server hint, not an exact row filter.

Predicates use ordinary nested R lists. The package serializes the list to the
JSON-string wire field required by Delta Sharing:

```{r, eval = FALSE}
active_orders <- list(
  op = "equal",
  children = list(
    list(op = "column", name = "is_active", valueType = "bool"),
    list(op = "literal", value = "true", valueType = "bool")
  )
)
orders$snapshot(predicate = active_orders)$to_data_frame()
```

## Read a change data feed

```{r, eval = FALSE}
orders$changes(starting_version = 120, ending_version = 125)$to_data_frame()
orders$changes(
  starting_timestamp = "2024-01-01T00:00:00Z",
  ending_timestamp = "2024-01-02T00:00:00Z"
)$to_data_frame()
```

## How reads work

For both Delta- and Parquet-format responses, R fetches the Query Table
response and writes a private local Delta log that Delta Kernel reads; Kernel
then fetches data directly from the pre-signed URLs. There is no second
downloader, and both formats flow through the same Kernel Arrow stream.
