---
title: "Performance and caching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance and caching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This guide describes the parts of a read that can be reused or tuned. Start
with `vignette("delta-sharing")` if you have not yet created a client and read a
table.

The most important distinction is between **staging files** and **reading
rows**. The package downloads the files selected by the sharing server before
it opens an Arrow stream. A lazy Arrow reader avoids eagerly collecting rows,
but it does not stream remote Parquet files directly from their signed URLs.

## The read lifecycle

Every snapshot or change read follows the same broad path:

1. R requests the relevant file actions from the Delta Sharing server.
2. Missing Parquet and deletion-vector files are downloaded into the table's
   session cache.
3. R writes a new local Delta log that points to the cached files.
4. Delta Kernel scans the local files and emits Arrow batches.
5. An eager materializer, when used, collects those batches into its result.

The stages have different reuse behavior:

| Work | Happens on another materializer call? | Reused from the session cache? |
|---|---:|---:|
| Sharing query and pagination | Yes | No |
| Download selected files | Only when missing or invalid | Yes |
| Build the local Delta log | Yes | No |
| Delta Kernel scan | Yes | No |
| Collect an eager result | Yes | No |

For snapshots, when the response format is not already known, a
`response_format = "auto"` read asks the metadata endpoint which format the
server selected. That choice is reused for the same table and client. Change
data feed reads request Delta format directly.

## What the cache contains

Each table has a deterministic cache directory under R's session temporary
directory. Its identity includes the profile endpoint and the table's share,
schema, and name. New handles for the same table therefore share cached files,
even when they use different download concurrency:

```{r, eval = FALSE}
client <- sharing_client("~/config.share")

orders <- client$table("sales.default.orders", concurrency = 4)
faster_downloads <- client$table("sales.default.orders", concurrency = 8)

identical(orders$cache_path, faster_downloads$cache_path)
#> TRUE
```

Parquet and deletion-vector files are named from immutable IDs supplied by the
sharing server. A later query can therefore reuse a file even when its signed
URL has rotated. If the server supplies a file size, the package checks that a
cached file still has that size before reusing it.

Downloads first go to temporary files beside the cache target. A completed file
is published with an atomic same-filesystem rename; failed or incomplete
downloads are not left as cache hits. Cached files are ordinary local copies of
the shared data, so the package requests user-only permissions for cache
directories and files.

## Cache lifetime

The cache lasts for the R session, not for the lifetime of one table handle.
It is normally removed with R's temporary directory when the session exits. It
is not a persistent, cross-session cache and it does not contain materialized
tibbles or Arrow tables.

`cache_path` is read-only but available for inspection:

```{r, eval = FALSE}
orders$cache_path
fs::dir_info(orders$cache_path)
```

An advanced user can remove that directory after all lazy readers and streams
for the table have been closed or released:

```{r, eval = FALSE}
# Do not do this while a lazy reader or stream is active.
if (fs::dir_exists(orders$cache_path)) {
  fs::dir_delete(orders$cache_path)
}
```

The next read recreates the directory and downloads any required files. A long
session that touches many large tables can retain substantial data until the
session ends, so `cache_path` is also useful when checking local disk usage.

## Cold and repeated reads

The first read of a set of files pays for their download. A later read can
reuse those files, but it still performs the Sharing request, creates a fresh
local log, scans the files with Delta Kernel, and materializes the result:

```{r, eval = FALSE}
snapshot <- orders$snapshot(
  columns = c("order_id", "status", "amount"),
  limit = 1000000
)

first <- system.time({
  first_result <- snapshot$to_tibble()
})

# Remove the first result only to benchmark another full materialization.
rm(first_result)
gc()

repeated <- system.time({
  repeated_result <- snapshot$to_tibble()
})
```

In application code, keep and reuse `first_result` when the same result is
needed more than once, rather than removing it as the benchmark does. The file
cache is most valuable when a later query overlaps the earlier query but needs
a fresh scan.

A fresh R session is the clearest definition of a cold read. The repository's
[`bench/snapshot.R`](https://github.com/zacdav-db/delta-sharing-r/blob/main/bench/snapshot.R)
script instead removes the table's file cache before each cold trial, then
measures the same query again with those files cached.

## Download concurrency

Set concurrency when creating a table handle:

```{r, eval = FALSE}
orders <- client$table(
  "sales.default.orders",
  concurrency = 8
)
```

The default is four. This value is the maximum number of missing remote assets
downloaded at once after all pages of the Sharing response have been collected.
It does not change API pagination, Delta Kernel scanning, Arrow batch size, or
a read that is already fully cached.

Higher concurrency is most likely to help when a query selects many files and
latency leaves the connection idle. It may do little when there are few files,
one large file already saturates the connection, or the provider throttles
requests. Treat four as the starting point and benchmark representative cold
reads before increasing it.

Interactive reads show progress for missing downloads. When every remote asset
has a known size, the display also includes the total bytes to download. Cache
hits do not appear as downloads.

## Shape the read before tuning it

Query options affect different stages:

- `columns` controls Delta Kernel projection and the shape of the materialized
  result. The current Sharing request does not send column projection, so it
  does not by itself reduce the size of already selected Parquet files.
- `limit` is sent to the server as a hint and is enforced exactly by Delta
  Kernel. A provider may use the hint to return fewer files, but this is not
  guaranteed.
- `predicate` is a best-effort server hint. It may improve file pruning, but it
  is not an exact row filter; apply an exact filter in the downstream consumer.
- `version` or `timestamp` changes which snapshot the server selects.

The default `response_format = "auto"` is normally the right choice. Setting
`"delta"` or `"parquet"` avoids the one-time negotiation request, but should be
done only when the provider supports the chosen format; it is not a general
performance switch.

## Choose a materializer

All materializers open a new read. They differ in how the Arrow batches are
consumed:

| Method | Collection behavior | Typical use |
|---|---|---|
| `to_tibble()` | Eager, in R memory | Ordinary R analysis |
| `to_data_frame()` | Eager, via the tibble path | Base data-frame consumers |
| `to_arrow()` | Eager, in Arrow memory | Repeated Arrow or DuckDB scans |
| `to_arrow_reader()` | Lazy after file staging | One-pass Arrow or DuckDB scan |
| `to_arrow_stream()` | Low-level and lazy after staging | Arrow C Stream consumers |

The Arrow table and reader methods require the optional `arrow` package. Close
an Arrow reader when it is no longer needed, and release a low-level stream if
its consumer does not take ownership.

For repeated DuckDB queries, registering an Arrow table or creating a temporary
DuckDB table avoids reopening the Delta Sharing read for every query. For one
pass over a large result, an Arrow reader avoids collecting an additional
in-memory Arrow table first.

## Arrow batch size

Every materializer accepts `batch_size`; the default is 65,536 rows:

```{r, eval = FALSE}
orders$snapshot()$to_tibble(batch_size = 65536)
```

Batch size controls how many rows the native reader emits in each Arrow batch.
It does not control the number or size of downloaded files. Very small batches
increase boundary and conversion overhead, while larger batches can increase
peak memory. The supported range is 1 to 1,000,000 rows, but the default is a
good starting point and should be changed only after measurement.

## Change data feeds

Snapshot and change readers for the same table use the same file cache.
Overlapping change ranges can therefore reuse immutable files already staged by
an earlier snapshot or change read. Every change request still builds a fresh
versioned local log and runs a new Delta Kernel change scan.

Large or overlapping historical ranges may add many files to the session cache.
The same concurrency and cache-lifetime guidance applies to those reads.

## A practical tuning order

1. Select only the columns and rows the consumer needs, while respecting the
   hint semantics above.
2. Reuse a materialized result when the same result is needed repeatedly.
3. Benefit from the session file cache for overlapping reads.
4. Compare a small number of download concurrency values on a representative
   cold read.
5. Choose the materializer that matches the consumer and memory budget.
6. Tune `batch_size` last, and only when measurements show batch overhead or
   memory pressure.

Measure end-to-end elapsed time and result size. Network throughput, latency,
provider behavior, file count, compression, schema width, and local hardware
all affect the result, so there is no universally optimal concurrency or batch
size.
